---
layout: post
title: "Mask a Raster using Threshold Values in R"
date:   2016-06-16
authors: [Leah A. Wasser, Kyla Dahlin]
instructors: [Leah, Naupaka]
time: "11:00"
contributors:
dateCreated:  2016-05-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rhdf5]
categories: [self-paced-tutorial]
mainTag: institute-day2
tags: [R, HDF5]
tutorialSeries: [institute-day2]
description: "This lesson covers how to mask, or how classify certain values in, a raster using threshold values."
code1: institute-materials/day2_tuesday/mask-raster.R
image:
  feature: 
  credit: 
  creditlink:
permalink: /R/mask-raster-threshold-R/
comments: false
---

## About

In this tutorial, we will walk through how to classify a raster file using
defined value ranges in R. 

First, let's load the required libraries.

```{r load-libraries, warning=FALSE, results='hide', message=FALSE}
# load libraries
library(raster)
library(rhdf5)
library(rgdal)

# be sure to set your working directory
# setwd("~/Documents/data/data-institute-2016") # Mac
# setwd("~/data/data-institute-2016")  # Windows

## import functions
# install devtools (only if you have not previously intalled it)
#install.packages("devtools")
# call devtools library
#library(devtools)

# install from github
#install_github("lwasser/neon-aop-package/neonAOP")
# call library
library(neonAOP)


#source("/Users/lwasser/Documents/GitHub/neon-aop-package/neonAOP/R/aop-data.R")


```

## Import LiDAR data

To begin, we will open the NEON LiDAR Digital Surface and Digital Terrain Models
(DSM and DTM) which are in Geotiff format.

```{r import-lidar }

# import aspect data from previous lesson
teak_nsAspect <- raster("outputs/TEAK/TEAK_nsAspect.tif")

# North facing slope = 1
# South facing slope = 2

# legend outside of the plot region
# make room for a legend
par(xpd = FALSE, mar=c(5.1, 4.1, 4.1, 4.5))

plot(teak_nsAspect, 
     col=c("white","blue","green"),
     main="North and South Facing Slopes \n Lower Teakettle",
     legend=F)

# allow legend to plot outside of bounds
par(xpd=TRUE)

legend((par()$usr[2] + 20), 4103300, # set xy legend location
       legend = c("North", "South"),
       fill = c("blue", "green"), 
       bty="n") # turn off border

```

## Mask Data

Once we have created a threhold classified raster, we can use it for different 
things. One application is to use it as an analysis mask for another dataset. 

Let's try to find all pixels that have an NDVI value >.6 and are north facing. 

```{r mask-data-ndvi }

# open NEON NDVI data
ndvi <- raster("NEONdata/D17-California/TEAK/2013/spectrometer/veg_index/NEON.D17.TEAK.DP2.20130614_100459_NDVI.tif")
ndvi

hist(ndvi,
     main="NDVI for Lower Teakettle Field Site")

# let's create a mask
ndvi[ndvi<.6] <- NA
plot(ndvi,
     main="NDVI > .6")

```


```{r mask-data }

n.face.aspect <- teak_nsAspect==1

# mask out only pixels that are north facing and NDVI >.6
nFacing.ndvi <- mask(n.face.aspect, ndvi)

plot(nFacing.ndvi,
     main="North Facing Locations \n NDVI > .6",
     legend=F)

```

## Export Classified Raster

```{r export-geotiff, eval=FALSE}

# export geotiff 
writeRaster(nFacing.ndvi,
            filename="outputs/TEAK/TEAK_n_ndvi6.tif",
            format="GTiff",
            options="COMPRESS=LZW",
            overwrite = TRUE,
            NAflag = -9999)


```