---
layout: post
title: "Calculate NDVI from NEON Hyperspectral Remote Sensing Data in R"
date:   2016-06-17
authors: [Leah A. Wasser, Kyla Dahlin]
instructors: [Leah, Naupaka]
time: "4:45"
contributors: [Edmund Hart]
dateCreated:  2016-05-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rhdf5]
categories: [self-paced-tutorial]
mainTag: institute-day1
tags: [R, HDF5]
tutorialSeries: [institute-day1]
description: "Intro to HDF5"
code1: institute-materials/day1_monday/calculate_ndvi_h5.R
image:
  feature:
  credit:
  creditlink:
permalink: /R/calculate-ndvi/
comments: false
---

First, let's load the required libraries.

```{r load-libraries, warning=FALSE, results='hide', message=FALSE}

# load libraries
library(raster)
library(rhdf5)
library(rgdal)

```

## Load Functions

This is a lot like loading a package in R. Except the functions for this package
are in an R script stored locally on our computers.

Once we have done the work to build our functions, we can perform routine tasks
over and over using those functions.

```{r load-functions }

source("/Users/lwasser/Documents/GitHub/neon-aop-package/neonAOP/R/aop-data.R")

```


## Calculate NDVI

Next we can use the `create_stack` function to create a raster stack of the
red and near-infrared bands that we need to calculate NDVI.

```{r create-NDVI }

# set working directory
setwd("~/Documents/data/1_data-institute-2016")
# Define the file name to be opened
f <- "Teakettle/may1_subset/spectrometer/Subset3NIS1_20130614_100459_atmcor.h5"

# define CRS
epsg=32611

# Calculate NDVI
# select bands to use in calculation (red, NIR)
ndvi_bands <- c(58, 90)

#create raster list and then a stack using those two bands
ndvi_stack <-  create_stack(ndvi_bands,
                            file = f,
                            epsg = epsg)

# calculate NDVI
NDVI <- function(x) {
	  (x[,2]-x[,1])/(x[,2]+x[,1])
}

ndvi_rast <- calc(ndvi_stack, NDVI)

# clear out plots
# dev.off(dev.list()["RStudioGD"])

plot(ndvi_rast,
     main="NDVI for the NEON TEAK Field Site")


```

## Export to GeoTiff

```{r export-ndvi, eval=FALSE}

# export as a gtif
writeRaster(ndvi_rast,
            file="ndvi_TEAK.tif",
            format="GTiff",
            overwrite=TRUE)

```

## Plot NDVI

```{r import-lidar}

DSM <- raster("Teakettle/may1_subset/lidar/Teak_lidarDSM.tif")  

slope <- terrain(DSM, opt='slope')
aspect <- terrain(DSM, opt='aspect')

# create hillshade
hill <- hillShade(slope, aspect, 40, 270)

plot(hill,
     col=grey(1:100/100),
     main="NDVI for the Teakettle Field site",
     legend=FALSE)

plot(ndvi_rast,
     add=TRUE,
     alpha=.3
     )
```
