---
layout: post
title: "Intro to working with Hyperspectral RS data in R -  H5"
date:   2016-06-20
authors: [Leah A. Wasser, Kyla Dahlin]
contributors: []
dateCreated:  2016-05-01
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rhdf5]
categories: [self-paced-tutorial]
mainTag: tabular-time-series
tags: [R, HDF5]
tutorialSeries: [institute-day1]
description: "Intro to HDF5"
code1: .R
image:
  feature: 
  credit: 
  creditlink:
permalink: /R/hdf5-R/
comments: false
---

First, let's load the required libraries.

```{r load-libraries, warning=FALSE, echo=FALSE}
# load libraries
library(raster)
library(rhdf5)
library(rgdal)

# set wd
setwd("~/Documents/data/1_spectrometerData/Teakettle/testSubset")
```

## Explore File
Next, define the file name and view the contents quickly.

```{r read-file}
# define the file name as an object
f <- "Subset1NIS1_20130614_095740_atmcor.h5"

# view the structure of the file
# reflectance is where the refl data are stored. 
# map info contains the proj information in WKT format
H5close()
h5ls(f, all = TRUE)

# let's grab some attribute information to use 
# here we have the CRS information that includes the 
# UPPER LEFT corner coordinate in UTM (meters)
mapInfo <- h5read(f,"map info", read.attributes = TRUE)
mapInfo

```

## View Wavelengths

Next, let's import some wavelength information.

```{r import-wavelength}
# import the center wavelength in um of each "band"
wavelengths<- h5read(f,"wavelength")

```

# Grab scale Factor
# notice the data ignore is a character.

```{r view-attr }

# r  get attributes for the Reflectance dataset
reflInfo <- h5readAttributes(f,"Reflectance")

# view the scale factor for the data
reflInfo$`Scale Factor`

# view the data ignore value
reflInfo$`data ignore value`

# the data ignore right now is a character (we are fixing this)
# you have to conver to num or int to ensure it is useful!


```

# Get shape of reflectance dataset

Let's 
read the data to grab the dimensions.
note: in a future version of the data we will have the dimensions as ATTRIBUTES
that you can automatically pull in

```{r import-reflectance }

# open the file for viewing
fid <- H5Fopen(f)
# open the reflectance dataset
did <- H5Dopen(fid, "Reflectance")
did

# grab the dimensions of the object
sid <- H5Dget_space(did)
dims <- H5Sget_simple_extent_dims(sid)$size

# take note that the data seem to come in ROTATED. wavelength is the 
# THIRD dimension rather than the first. Columns are the FIRST dimension, then rows.
# close everything
H5Sclose(sid)
H5Dclose(did)
H5Fclose(fid)

```

Once we know the dimensions of the data, we can more efficient slice out chunks of it
Let's first grab all of the data for band34

```{r read-refl-data }
# if you get an error with the file being "open" just use the generic h5 close below
# when we are done with our attributes you can skip all of this nonsense :)
H5close()
# Extract or "slice" data for band 34 from the HDF5 file
b56<- h5read(f,"Reflectance", index=list(1:dims[1],1:dims[2],56))
# note the data come in as an array
class(b56)

```

Next, we will convert the data to a matrix and then to a raster.
We don't need an array because our data are only 2 dimensions at this point (1 single band).

```{r view-data }
# Convert from array to matrix so we can plot and convert to a raster
b56 <- b56[,,1]

# plot the data
# what happens when we plot?
image(b56)
# looks like we need to force a stretch
image(log(b56), main="band 56 with log transformation")
# view distribution
hist(b56)

```

Ok so now we need to do a few things

1. set the no data value (15000)
2. scale the data by 10000

```{r no-data-scale }

noDataVal <- as.integer(reflInfo$`data ignore value`)
# set all values = 15,000 to NA
b56[b56 == noDataVal] <- NA

# lets use our scale factor first
scaleFactor <- reflInfo$`Scale Factor`
b56 <- b56/scaleFactor

hist(b56, main="distribution with NoData Value considered\nData scaled")
```


Note -- there are still issues with values over 1. I need to followup with
AOP about how to handle these. 
```{r transpose-data }
# Because the data import column, row but we require row, column in R, 
# We need to transpose x and y values in order for our final image to plot properly
b56<-t(b56)
image(log(b56), main="Transposed image")

```

Next, we need to create the spatial extent required to position the raster in space. 
```{r}
# so we can extract the lower left hand corner coordinates.
# the numbers as position 4 and 5 are the UPPER LEFT CORNER (x,y)
mapInfo<-unlist(strsplit(mapInfo, ","))

# grab the XY left corner coordinate
xMin <- as.numeric(mapInfo[4])
# ensure the format is numeric
yMax <- as.numeric(mapInfo[5])
# we can get the x and y resolution from this string too
res <- c(mapInfo[2],mapInfo[3])
res <- as.numeric(res)

#  finally calculate the xMAX value and the YMIN value
# we grabbed the dimensions above. the xmax is the left corner + number of columns* resolution
xMax <- xMin + (dims[1]*res[1])
yMin <- yMax - (dims[2]*res[2])

# also note that x and y res are the same (1 meter)
# Now, define the raster extent
# define the extent (left, right, top, bottom)
rasExt <- extent(xMin, xMax,yMin,yMax)

# now we can create a raster and assign it it's spatial extent
b56r <- raster(b56,
               crs=CRS("+init=epsg:32611"))
# assign CRS
extent(b56r) <- rasExt

# note that the raster flips the data once again.
# i still need to sort out how to explain this but it has to do with how
# the raster format nativly is stored.
b56r
plot(b56r, main="Raster for Teakettle - B56")

```


