---
layout: post
title: "Subset HDF5 file in R"
date:   2016-06-17
authors: [Leah A. Wasser]
instructors: []
contributors: []
time: "3:30 pm"
dateCreated:  2016-05-10
lastModified: `r format(Sys.time(), "%Y-%m-%d")`
packagesLibraries: [rhdf5]
categories: [self-paced-tutorial]
mainTag: institute-day1
tags: [R, HDF5]
tutorialSeries: [institute-day1]
description: "Learn how to subset an H5 file in R."
code1: subset-h5-file-R.R
image:
  feature: 
  credit: 
  creditlink:
permalink: /R/subset-hdf5-R/
comments: false
---


<div id="objectives">
<strong>R Skill Level:</strong> Intermediate

<h3>Goals / Objectives</h3>
After completing this activity, you will:
<ol>
<li>Understand how to extract and plot spectra from an HDF5 file.</li>
<li>Know how to work with groups and datasets within an HDF5 file.</li>
</ol>

</div> 

## Getting Started

In this tutorial, we will extract a single-pixel's worth of reflectance values 
from an HDF5 file and plot a spectral profile for that pixel.

```{r call-libraries, results="hide" }

#first call required libraries
library(rhdf5)
library(raster)
library(plyr)
library(rgeos)
library(rgdal)
library(ggplot2)

# be sure to set your working directory
setwd("~/Documents/data/1_data-institute-2016")

```

## Import H5 Functions

In this scenario, we have built a suite of FUNCTIONS that will allow us to quickly
open and read NEON hyperspectral imagery data from an `hdf5` file. We can import
a suite of functions from an `.R` file using the `source` function. 

```{r import-h5-functions }

# your file will be in your working directory! This one happens to be in a diff dir
# than our data

source("/Users/lwasser/Documents/GitHub/neon-data-institute-2016/_posts/institute-materials/day1_monday/import-HSIH5-functions.R")

```

## Open H5 File

Next, let's define a few variables, that we will need to access the H5 file.

```{r open-H5-file, results='hide'}

# Define the file name to be opened
f <- "Teakettle/may1_subset/spectrometer/Subset3NIS1_20130614_100459_atmcor.h5"

# Look at the HDF5 file structure 
h5ls(f,all=T) 

# define the CRS in EPGS format for the file
epsg <- 32611

```

## Read Wavelength Values

Next, let's read in the wavelength center associated with each band in the HDF5 
file. 


```{r read-band-wavelengths }

#read in the wavelength information from the HDF5 file
wavelengths<- h5read(f,"wavelength")
# convert wavelength to nanometers (nm)
# NOTE: this is optional!
wavelengths <- wavelengths*1000

```




## Get Subset
Next, we might want to extract just a subset of our data to pull a spectral
signature. For example a plot boundary.

```{r extract-subset}

# get of H5 file map tie point
h5.ext <- create_extent(f)

# turn the H5 extent into a polygon to check overlap
h5.ext.poly <- as(extent(h5.ext), "SpatialPolygons")

# open file clipping extent
clip.extent <- readOGR("Teakettle", "teak_plot")

# assign crs to h5 extent
# paste0("+init=epsg:", epsg) -- so it will be better to use the proj string here

crs(h5.ext.poly) <- CRS("+proj=utm +zone=11 +datum=WGS84 +units=m +no_defs +ellps=WGS84 +towgs84=0,0,0")
  

# ensure the two extents overlap
gIntersects(h5.ext.poly, clip.extent)

# if they overlap, then calculate the extent
# this doesn't currently account for pixel size at all
# also these values need to be ROUNDED

yscale <- 1
xscale <- 1

# define index extent
# xmin.index, xmax.index, ymin.index,ymax.index
# all units will be rounded which means the pixel must occupy a majority (.5 or greater) 
# within the clipping extent
index.bounds <- calculate_index_extent(extent(clip.extent), 
                                       h5.ext)

# open a band that is subsetted using the clipping extent
b58_clipped <- open_band(fileName=f,
          bandNum=58,
          epsg=32611,
          subsetData = TRUE,
          dims=index.bounds)

# plot clipped bands
plot(b58_clipped,
     main="Band 58 Clipped")

```

## Run subset over many bands

```{r, open-many-bands }

# create  alist of the bands
bands <- list(19,34,58)
# within the clipping extent
index.bounds <- calculate_index_extent(extent(clip.extent), 
                                            h5.ext)
rgbRast.clip <- create_stack(file=f,
                         bands=bands,
                         epsg=epsg,
                         subset=TRUE,
                         dims=index.bounds)

plotRGB(rgbRast.clip,
        stretch="lin")

```

## Unclipped Data
And here are the same data not subsetted but with the subset region overlayed
on top.

```{r}

rgbRast <- create_stack(file=f,
                         bands=bands,
                         epsg=epsg,
                         subset=FALSE)

plotRGB(rgbRast,
        stretch="lin")
plot(clip.extent, 
     add=T,
     border="yellow",
     lwd=3)

```

## View Spectra Within Clipped Raster

```{r subset-h5-file}

# array containing the index dimensions to slice
H5close()
subset.h5 <- h5read(f, "Reflectance",
                    index=list(index.bounds[1]:index.bounds[2], index.bounds[3]:index.bounds[4], 1:426)) # the column, row 

final.spectra <- data.frame(apply(subset.h5,
              MARGIN = c(3), # take the mean value for each z value 
              mean)) # grab the mean value in the z dimension
final.spectra$wavelength <- wavelengths

# scale the data
names(final.spectra) <- c("reflectance","wavelength")
final.spectra$reflectance <- final.spectra$reflectance / 10000


# plot the data
qplot(x=final.spectra$wavelength,
      y=final.spectra$reflectance,
      xlab="Wavelength (nm)",
      ylab="Reflectance",
      main="Mean Spectral Signature for Clip Region")

```


Now we've plotted a spectral signature. However, the values are averaged over all pixels in our
AOI. We may just want to explore the spectral of particular, thresholded object.

to do this

* i need to find the new tie point which will be a result of any averageing done above
* then i can spatially located this and turn it into a raster potentially?
* then i can create a MASK of just pixels with a particular raster value.

Let's give it a go

```{r import-tif }
# create a list of bands
bands <- c(60,83)
index.bounds <- calculate_index_extent(extent(clip.extent), 
                                            h5.ext)


ndvi.stack <- create_stack(f, 
                           bands, 
                           epsg=32611,
                           subset=TRUE,
                           dims=index.bounds)

# calculate NDVI
ndvi <- (ndvi.stack[[2]]-ndvi.stack[[1]]) / (ndvi.stack[[2]]+ndvi.stack[[1]])
names(ndvi) <- "Teak_hsiNDVI"

# let's test this out
plot(ndvi)

# let's create a mask
ndvi[ndvi<.6] <- NA
plot(ndvi)

```



```{r function-extract-spectra }

## FUNCTION - Extract Average Reflectance
#'
#' This function calculates an index based subset to slice out data from an H5 file
#' using an input spatial extent. It returns a rasterStack object of bands. 
#' @param aRaster REQUIRED. a raster within an H5 file that you wish to get an mean, max, min value from
#' @param aMask REQUIRED. a raster of the same extent with NA values for areas that you don't 
#' want to calculate the mean, min or max values from. 
#' @param aFun REQUIRED. the function that you wish to use on the raster, mean, max, min etc
#' @keywords hdf5, extent
#' @export
#' @examples
#' extract_av_refl(aRaster, aMask, aFun=mean)
#' 
extract_av_refl <- function(aRaster, aMask=NULL, aFun=mean){
  # mask band
  if(!is.null(aMask)){
  aRaster <- mask(aRaster, aMask) }
  # geat mean
  a.band.mean <- cellStats(aRaster, 
                         aFun, 
                         na.rm=TRUE)
  return(a.band.mean)
} 

## FUNCTION - Extract Average Reflectance
#'
#' This function calculates an index based subset to slice out data from an H5 file
#' using an input spatial extent. It returns a rasterStack object of bands. 
#' @param fileName REQUIRED. The path to the h5 file that you want to open.
#' @param bandNum REQUIRED. The band number that you wish to open, default = 1 
#' @param epsg the epsg code for the CRS that the data are in.
#' @param subsetData, a boolean object. default is FALSE. If set to true, then
#' ... subset a slice out from the h5 file. otherwise take the entire xy extent.
#' @param dims, an optional object used if subsetData = TRUE that specifies the 
#' index extent to slice from the h5 file
#' @param mask a raster containg NA values for areas that should not be included in the 
#' output summary statistic. 
#' @param fun the summary statistic that you wish to run on the data  (e.g. mean, max, min)
#' default = mean
#' @keywords hdf5, extent
#' @export
#' @examples
#' get_spectra(fileName, bandNum)

get_spectra <- function(fileName, bandNum=1, 
                           epsg=32611, subset=TRUE,
                           dims=NULL, mask=NULL, fun=mean){
  # open a band
  a.raster <- open_band(fileName, bandNum, 
                           epsg, subset,
                           dims)
  # extract a particular spectral value (min, max, mean from the raster)
  refl <- extract_av_refl(a.raster, mask, aFun = fun)
  return(refl)
}
```

## Plot Spectra

```{r plot-spectra-unmasked }


# if you run get_spectra on it's own, it returns a min, max or mean value for a raster
get_spectra(f, bandNum=1, epsg=32611, subset=TRUE,
                           dims=index.bounds, mask=ndvi, fun=mean)

# provide a list of bands that you wish to extract summary values for
bands <- (1:426)


spectra_unmasked <- lapply(bands, FUN = get_spectra,
              fileName=f, epsg=32611,
              subset=TRUE,
              dims=index.bounds, fun=mean)


# reformat the output list
spectra_unmasked <- unlist(spectra_unmasked)

# plot spectra
plot(spectra_unmasked,
     main="Spectra for all pixels",
     xlab="Band",
     ylab="Reflectance")

# run get_spectra for each band to get an average spectral signature
# because we've specified a mask, it will only return values for pixels that are not
# in masked areas
spectra <- lapply(bands, FUN = get_spectra,
              fileName=f, epsg=32611,
              subset=TRUE,
              dims=index.bounds, 
              mask=ndvi, fun=mean)

# reformat the output list
spectra <- unlist(spectra)

# plot spectra
plot(spectra,
     main="Spectra for just green pixels",
     xlab="Band",
     ylab="Reflectance")

```


